\documentclass[a4paper,12pt,reqno]{amsart}
\usepackage[utf8]{inputenc}
\usepackage[top=25truemm,bottom=25truemm,left=20truemm,right=20truemm]{geometry}

\usepackage{amsmath, amssymb, amsfonts, amscd, amsthm, mathtools}
\newtheorem{lemma}{Lemma}

\title{A classification of congruences in the category of sets and relations}
\author{Daniel Schepler}
\date{\today}

\begin{document}
\maketitle

We will give a classification of congruences $i : E \hookrightarrow X \times X$ in the category $\mathbf{Rel}$ of sets and relations.  Recall that products in $\mathbf{Rel}$ are actually given by disjoint unions at the level of underlying sets.  Also, recall that a morphism $R : X \to Y$ in $\mathbf{Rel}$ is a monomorphism if and only if $R_* : P(X) \to P(Y), S \mapsto \{ y\in Y \mid \exists x\in S, (x, y) \in R \}$ is injective, and it is an isomorphism if and only if $R$ is the graph of a bijection $X \to Y$ in $\mathbf{Set}$.

In particular, we get $i_* : P(E) \to P(X + X) \simeq P(X) \times P(X)$ which must be injective.  It must also preserve arbitrary unions and in particular be inclusion-preserving.  From the assumption that $i$ is a congruence, since the functor $(P, {-}_*) : \mathbf{Rel} \to \mathbf{Set}$ is representable by the object 1, we see that $i_*$ must have image given by an equivalence relation $\sim$ on $P(X)$.  Note also that since $i_*$ preserves arbitrary unions, we must have that $\sim$ respects arbitrary unions.

Since the symmetry morphism $s : E \to E$ satisfies $s^2 = \operatorname{id}$, it must be the graph of an involution $s_0$ on $E$, where $s_0$ is a morphism in $\mathbf{Set}$.

{\bf Claim 1:} Consider the reflexivity morphism $r : X \to E$. For each $x \in X$, $r_*(\{ x \})$ must be either a singleton or a pair.

To see this, $r_*(\{ x \})$ certainly cannot be empty, or we would have $i_*(r_*(\{ x \})) = (\varnothing, \varnothing) \ne (\{ x \}, \{ x \})$.  On the other hand, if we had at least three distinct elements $a,b,c \in r_*(\{ x \})$, then $i_*(\varnothing), i_*(\{ a \}), i_*(\{ a, b \}), i_*(\{ a, b, c \})$ would have to be a chain of four distinct subsets of $i_*(r_*(\{ x \})) = (\{ x \}, \{ x \})$, which is impossible.

{\bf Claim 2:} For each $x \in X$, if $r_*(\{ x \}) = \{ e, f \}$ with $e \ne f$, then $s_0(e) = f$ and $s_0(f) = e$; and we have either that $i_*(\{e\}) = (\varnothing, \{ x \})$ and $i_*(\{f\}) = (\{ x \}, \varnothing)$, or vice versa.  Otherwise, if $r_*(\{ x \}) = \{ e \}$, then $s_0(e) = e$; and $i_*(\{ e \}) = (\{ x \}, \{ x \})$.

To see this, if $r_*(\{ x \}) = \{ e, f \}$ with $e\ne f$, then we must have $\varnothing = i_*(\varnothing) \subsetneqq i_*(\{ e \}) \subsetneqq i_*(\{ e, f \}) = (\{ x \}, \{ x \})$, which means that $i_*(\{ e \})$ is equal to either $(\{ x \}, \varnothing)$ or $(\varnothing, \{ x \})$.  By the same token, $i_*(\{ f \})$ is also one of those choices, and $i_*(\{ e \}) \ne i_*(\{ f \})$.  Since $i_*(\{ s_0(e) \}) = i_*(s_*(\{ e \})) = i_*(\{ f \})$ in either choice of order, by injectivity of $i_*$ we must have that $s_0(e) = f$.  Similarly, $s_0(f) = e$.  Otherwise, if $r_*(\{ x \}) = \{ e \}$ is a singleton, $i_*(\{ s_0(e) \}) = i_*(s_*(\{ e \})) = i_*(\{ e \})$, so injectivity of $i_*$ gives $s_0(e) = e$.

{\bf Claim 3:} For $x\in X$, suppose that $r_*(\{ x \}) = \{ e \}$.  Then $\{ x \} \not\sim \varnothing$.

If we did have $\{ x \} \sim \varnothing$, then there would have to be some subset of $E$ which maps to $(\varnothing, \{ x \})$.  Similarly to previous steps, injectivity of $i_*$ implies the subset is nonempty, and if the subset had two distinct elements $a,b$, then we would have $i_*(\varnothing) \subsetneqq i_*(\{a\}) \subsetneqq i_*(\{a,b\}) \subseteq (\varnothing, \{ x \})$, giving a contradiction.  Therefore, this subset of $E$ would have to be a singleton $\{ f_1 \}$.  Similarly, $(\{ x \}, \varnothing)$ would have to be the image of a singleton $\{ f_2 \}$ with $f_1 \ne f_2$.  But then $i_*(\{ f_1, f_2 \}) = (\{ x \}, \{ x \}) = i_*(\{ e \})$, contradicting the injectivity of $i_*$.

{\bf Claim 4:} For any $e \in E$, $i_*(\{ e \})$ is of one of the forms $(\{ x \}, \varnothing)$, $(\varnothing, \{ x \})$, or $(\{ x \}, \{ x \})$.  In the first two cases, $s_0(e) \ne e$ and $\{ x \} \sim \varnothing$; and in the third case, $s_0(e) = e$ and $\{ x \} \not\sim \varnothing$.

To show this, let $e\in E$, and suppose that $i_*(\{ e \}) = (S, T)$.  Then $i_*(\{ s_0(e) \}) = (T, S)$, so $i_*(\{ e, s_0(e) \}) = (S \cup T, S\cup T) = i_*(\bigcup_{x\in S\cup T} r_*(\{ x \}))$.  It follows that $\bigcup_{x\in S\cup T} r_*(\{ x \}) = \{ e, s_0(e) \}$.  Therefore, $e \in r_*(\{ x \})$ for some $x \in X$, and by claims 2 and 3, we get the desired conclusion.

{\bf Claim 5:} For any $S, T \in P(X)$, $S \sim T$ if and only if $S\cap A = T\cap A$, where $A$ is the set of $x\in X$ such that $r_*(\{ x \})$ is a singleton.

For the forward direction, suppose $(S, T) = i_*(U)$ for $U \subseteq E$.  Then $(S, T) = \bigcup_{e\in U} i_*(\{ e \})$; $i_*(\{ e \})$ satisfies the relation of having equal intersections with $A$ for each $e\in E$ in any case from claim 4; and this relation respects unions.  For the reverse implication, whenever $x\notin A$, we have $\{ x \} \sim \varnothing$.  Therefore, since $S = (S \cap A) \cup \bigcup_{x\in S \setminus A} \{ x \}$, we must have $S \sim (S \cap A) \cup \bigcup_{x\in S \setminus A} \varnothing = S \cap A$.  Similarly, $T \sim T \cap A$, so if $S\cap A = T\cap A$, then $S \sim T$.

{\bf Claim 6:} Let $E'$ be the set pushout $X +_A X$ with inclusion maps $i_1, i_2 : X \to E'$. Define a morphism $E' \to X \times X$ in $\mathbf{Rel}$ given by the inverse relation of the graph $i_1 + i_2 : X + X \to E'$. Then the congruences $E$ and $E'$ are equivalent.

We can define a bijection $E \to E'$ as follows: let $e \in E$.  If $i_*(\{ e \}) = (\{ x \}, \varnothing)$ with $x\notin A$, then send $e \mapsto i_1(x)$; if $i_*(\{ e \}) = (\varnothing, \{ x \})$ with $x\notin A$, then send $e \mapsto i_2(x)$; and if $i_*(\{ e \}) = (\{ x \}, \{ x \})$ with $x \in A$, then send $e \mapsto i_1(x) = i_2(x)$.  Transferring the congruence to $E'$, we see that it is exactly of the given form.

~

Now, without loss of generality we will assume from now on that the congruence is of exactly this form.

With this classification, we can show that $A$ is a quotient for the congruence, with quotient morphism $\Delta_A \subseteq A \times A \subseteq X \times A$.  In fact, we can define a morphism $\Delta_A^\circ : A \to X$ as the inverse relation to $\Delta_A$, and a morphism $X \to E$ as the graph of $i_1$.  It is then straightforward to verify that these maps together make $A$ a split coequalizer of $p_1 \circ i$ and $p_2 \circ i$.

We can also conclude that $E$ is the kernel pair of this quotient, by the general result below.

\begin{lemma}
	Suppose we have a congruence $f, g : E \rightrightarrows X$ with a contractible coequalizer
	$$ E \overset{f}{\underset{g}{\rightrightarrows}} X \overset{e}{\rightarrow} Q $$
	with maps in the reverse direction $s : Q \to X$ and $t : X \to E$.  Then $E$ is the kernel pair of this quotient, i.e.~ we have a cartesian square
    $$
    \begin{CD}
	    E @> f >> X \\
	    @V g VV @V e VV \\
	    X @> e >> A.
    \end{CD}
    $$
    \begin{proof}
    	Suppose we have generalized elements $x_1, x_2 : U \to X$ with $e x_1 = e x_2$.  Then $f t x_1 = x_1$ and $g t x_1 = s e x_1$, so the pair $x_1, s e x_1$ factors through $E$.  Similarly, the pair $x_2, s e x_2$ factors through $E$.  However, by the assumption, we also have $s e x_1 = s e x_2$.  Therefore, since $E$ is a congruence, we conclude $x_1, x_2$ factors through $E$.
    \end{proof}
\end{lemma}

\end{document}
